---
title: "DATA612_Final_Project_Southwood"
format: html
editor: visual
---

## Loading in data and manipulation

```{r}
# load packages 
library(tidyverse)
library(ggplot2)
```

```{r}
# load data from https://www.kaggle.com/datasets/warcoder/earthquake-dataset/data
data <- read_csv("earthquake_1995-2023.csv")

# look at data 
head(data)
```

```{r}
# edit date time data to only have date 
data$date_only <- as.Date(data$date_time, format = "%d-%m-%Y %H:%M")

# view new column 
head(data)
```

## Preliminary plots

```{r}
# exploring depth and magnitude 
ggplot(data, aes(x = depth, y = magnitude, color = alert )) + geom_point()
```

```{r}
# exploring gap, magType, and date 
ggplot(data, aes(x = date_only, y = gap, color = magType)) + geom_point()
```

## Final plots

Scatter Plot: Magnitude vs. Significance, colored by alert

```{r}

# drop rows with NA in alert column 
data_dropNA_alert <- data |> drop_na(alert)

# plot scatterplot  
ggplot(data_dropNA_alert, aes(x = magnitude, y = sig, fill = alert)) + 
  geom_point(shape = 21, color = "black", size = 3, stroke = 0.5) +
  scale_fill_manual(values = c("green", "yellow", "orange", "red")) +
  labs(title = "Earthquake Magnitude vs Significance",
       x = "Magnitude",
       y = "Significance",
       fill = "Alert") +
  theme_bw()

```

 Boxplot: Magnitude by alert level

```{r}

# plot boxplot 
ggplot(data_dropNA_alert, aes(x = alert, y = magnitude, fill = alert)) +
  geom_boxplot() +
  scale_fill_manual(values = c("green", "yellow", "orange", "red")) +
  labs(title = "Earthquake Magnitude by Alert Level",
       x = "Alert",
       y = "Magnitude") + 
  theme_bw() + 
  theme(legend.position = "none")

```

 Histogram: Magnitude distribution over time (faceted by 5-year period) and counts

```{r}

# add column that sorts dates into five year periods 
data<- data |> 
      mutate(five_year_period = factor(floor(as.numeric(format(date_only, "%Y")) / 5) * 5))

# plot histogram 
ggplot(data, aes(x = magnitude)) +
  geom_histogram(binwidth = 0.5, fill = "pink", color = "black") +
  facet_wrap(~ five_year_period) +
  labs(title = "Magnitude Distribution by 5-Year Period",
       x = "Magnitude",
       y = "Frequency") +
  theme_bw()

```

```{r}

# count number of earthqaukes in each five year period 
earthquake_counts <- data |>
  group_by(five_year_period) |>
  summarise(n_earthquakes = n()) |>
  arrange(five_year_period)

# view counts 
earthquake_counts
```

 Histogram: Magnitude distribution by Tsunami and counts

```{r}

# plot histogram 
ggplot(data, aes(x = magnitude, fill = factor(tsunami))) +
  geom_histogram(binwidth = 0.5, color = "black") +
  facet_wrap(~ factor(tsunami, labels = c("Land", "Oceanic"))) +
  scale_fill_manual(values = c("grey", "lightblue")) +
  labs(title = "Oceanic vs Land Earthquake Distribution",
       x = "Magnitude",
       y = "Frequency") +
  theme_bw() + 
  theme(legend.position = "none")


```

```{r}

# count number of earthqaukes on land and in the ocean  
tsunami_counts <- data |>
  group_by(tsunami) |>
  summarise(n_earthquakes = n()) |>
  arrange(tsunami)

# view counts 
tsunami_counts
```

 Boxplot: Significance by alert level (NOT included in final presentation)

```{r}

# plot box plot by significance 
ggplot(data_dropNA_alert, aes(x = alert, y = sig, fill = alert)) + geom_boxplot() + scale_fill_manual(values = c("green", "yellow", "orange", "red")) +
  labs(title = "Earthquake Significance by Alert Level",
       x = "Alert",
       y = "Significance",
       color = "Alert") +
 theme_bw()
```
